---
title: "In-class Exercise 04 "
author: "Jezelei Manangan"
date: "September 9, 2024"
date-modified: "last-modified"
execute: 
  eval: true
  echo: true
  freeze: true
---

# 1.0 Installing Packages

```{r}
pacman::p_load(arrow,lubridate,tidyverse,tmap,sf)
```

```{r}
#| eval: false
install.packages("maptools", repos = "https://packagemanager.posit.co/cran/2023-10-13")
```

```{r}
acled_sf<- read_csv("data/aspatial/ACLED_Myanmar.csv") %>% 
  #convert into sf
  st_as_sf(coords = c("longitude", "latitude"),
           crs=4326) %>% 
  st_transform(crs = 32647) %>% 
  mutate(event_date = dmy(event_date))
```

Look at the date-field specifically, without conversion, the event_date is converted into character instead of retaining DD-MMM-YYYY format. To fix this, add the mutate function and transform into sf

# 2.0 Importing Datasets to R Environment

In this exercise, we will use the following datasets:

## 2.1 Importing Geospatial Data

In this section, [*st_read()*](https://r-spatial.github.io/sf/reference/st_read.html) of **sf** package will be used to import the three geospatial data sets mentioned in previous section into R environment./r

```{r}
childcare_sf <- st_read("data/aspatial/child-care-services-geojson.geojson")
```

```{r}
mpsz_sf <- st_read(dsn = "data/geospatial", layer = "MP14_SUBZONE_WEB_PL")
```

# 3.0 GeoSpatial Data Wrangling

## 3.1 Assigning Standard CRS

```{r}
childcare_sf <- st_transform(childcare_sf, crs = 3414)
mpsz_sf <- st_transform(mpsz_sf, crs = 3414)
```

```{r}
sg_sf <- mpsz_sf %>% 
  st_union()
```

```{r}
plot(sg_sf)
```

```{r}
tmap_mode("plot")
acled_sf %>% 
  filter(year == 2023 | event_type == "Political Violence") %>% 
  tm_shape() + 
  tm_dots() 
```

```{r}
tmap_mode('plot')
```

# 4.0 

# 5.0
