---
title: "Hands-on Exercise 11"
date: "October 27, 2024"
date-modified: "last-modified"
execute: 
  eval: true
  echo: true
  freeze: true
---

# Overview

**Geographically weighted regression (GWR)** is a spatial statistical technique that takes non-stationary variables into consideration (e.g., climate; demographic factors; physical environment characteristics) and models the local relationships between these independent variables and an outcome of interest (also known as dependent variable). In this hands-on exercise, you will learn how to build [hedonic pricing](https://www.investopedia.com/terms/h/hedonicpricing.asp) models by using GWR methods. The dependent variable is the resale prices of condominium in 2015. The independent variables are divided into either structural and locational.

## The Data

Two data sets will be used in this model building exercise, they are:

-   URA Master Plan subzone boundary in shapefile format (i.e. *MP14_SUBZONE_WEB_PL*)

-   condo_resale_2015 in csv format (i.e. *condo_resale_2015.csv*)

## Loading the Data

The code chunks below installs and launches the necessary R packages into R environment.

```{r}
pacman::p_load(olsrr, corrplot, ggpubr, sf, spdep, GWmodel, tmap, tidyverse, gtsummary, ggstatsplot, performance)
```

```{r}
condo_resale <- read_csv("data/aspatial/Condo_resale_2015.csv")
mpsz <- st_read(dsn = "data/geospatial", layer = "MP14_SUBZONE_WEB_PL")
condo_resale_sf <- st_as_sf(condo_resale,
                            coords = c("LONGITUDE", "LATITUDE"),
                            crs=4326) %>%
  st_transform(crs=3414)
```

# Correlation Analysis

```{r}
ggcorrmat(condo_resale[, 5:23])
```

## Building a Hedonic Pricing Model using Multiple Linear Regression Method

```{r}
condo_mlr <- lm(formula = SELLING_PRICE ~ AREA_SQM + AGE  + 
                        PROX_CBD + PROX_CHILDCARE + PROX_ELDERLYCARE    + 
                        PROX_URA_GROWTH_AREA + PROX_MRT + PROX_PARK + 
                        PROX_PRIMARY_SCH + PROX_TOP_PRIMARY_SCH + PROX_SHOPPING_MALL   +
                        PROX_BUS_STOP + PROX_SUPERMARKET + 
                        NO_Of_UNITS + FAMILY_FRIENDLY +
                        FREEHOLD + LEASEHOLD_99YR, 
                data= condo_resale_sf)

summary(condo_mlr)
```

Front - Dependent Variable

-   Matrix does not work here, needs to be a data frame input

# Generating tidy linear regression report

```{r}
ols_regress(condo_mlr)
```

# Variable Selection

```{r}
condo_fw_mlr <- ols_step_forward_p(
  condo_mlr, 
  p_val = 0.05, 
  details= FALSE) 
condo_fw_mlr
```

Any independent variable we add in, it will compare the variable at every iteration.

# Building Backward Stepwise MLR

```{r}
condo_bw_mlr <- ols_step_backward_p(
  condo_mlr, 
  p_val = 0.05, 
  details= FALSE) 
condo_bw_mlr
```

# Building Stepwise Multiple Linear Regression

```{r}
condo_sb_mlr <- ols_step_both_p(
  condo_mlr, 
  p_val = 0.05, 
  details= FALSE) 
condo_sb_mlr
```

# Model Selection

```{r}
metric <- compare_performance(condo_mlr, 
                              condo_fw_mlr$model, 
                              condo_bw_mlr$model, 
                              condo_sb_mlr$model)
```

```{r}
metric$Name <- gsub(".*\\\\[a-zA-Z0-9_]+\\\\,\\\\model\\\\.*","\\1", metric$Name)
```

```{r}
#plot(metric)
```

# Checking for Multicollinearity

```{r}
check_collinearity(condo_sb_mlr$model)
```

```{r fig.width=12, fig.height=6}
plot(check_collinearity(condo_sb_mlr$model)) + 
  theme(axis.text.x = element_text(
    angle = 45, hjust = 1
  ))
```

# Linearity Assumption Test

```{r}
#out <- plot(check_model(condo_sb_mlr$model, panel=FALSE))

#out[[2]]
```

# Checking for outliers

```{r}
#plot(check_outliers(condo_sb_mlr$model, 
                    #method = "cook"))
```
