---
title: "In-class Exercise 13"
date: "November 11, 2024"
date-modified: "last-modified"
execute: 
  eval: true
  echo: true
  freeze: true
---

# Overview

Predictive modelling uses statistical learning or machine learning techniques to predict outcomes. By and large, the event one wants to predict is in the future. However, a set of known outcome and predictors (also known as variables) will be used to calibrate the predictive models.R Packages

# Data

## R Packages

```{r}
pacman::p_load(sf, spdep, GWmodel, SpatialML, 
               tmap, rsample, yardstick, tidyverse, knitr, kableExtra, spatialRF)
```

## Preparing Data

1.  Data Import

```{r}
mdata <- read_rds("data/rds/mdata.rds")
```

2.  Data Sampling

Calibrating predictive models are computationally intensive, especially when random forest method is used. For quick prototyping, a 10% sample will be selected at random from the data using the code chunk below.

-   Consider RStudio Timeout – if there are huge data points and the calibration will take a very long time.

-   Setting a sample size such as 1500 in this case

```{r}
set.seed(1234)
HDB_sample <- mdata %>% sample_n(1500)
```

3.  Checking of Overlapping Point

```{r}
overlapping_points <- HDB_sample %>% 
  mutate(overlap = lengths(st_equals(., .))>1)

summary(overlapping_points$overlap)
```

3.  Spatial Jitter

In the code chunk below, `st_jitter()` of sf package is used to move the point features by 5m to avoid overlapping point features.

```{r}
HDB_sample <- HDB_sample %>% 
  st_jitter(amount=5)
```

-   Don't use decimal places because there are truncations

-   If the unit is in meters, do in meters

# Data Sampling

The entire dataset is split into testing and training - 65% and 35% respectively using `initial_split()` of **rsample** package.

```{r}
set.seed(1234)
resale_split <- initial_split(HDB_sample, prop = 6.67/10,)

train_data <- training(resale_split)
test_data <- testing(resale_split)
```

We save the data in RDS:

```{r}
write_rds(train_data, "data/rds/train_data.rds")
write_rds(test_data, "data/rds/test_data.rds")
```

## Multicollinearity Check

In order to avoid multicollinearity, we use `ggcorrmat()` of **ggstatplot** to plot a correlation matrix to check if there are pairs of highly correlated independent variables:

```{r fig.width=12, fig.height=12}
mdata_nogeo <- mdata %>% 
  st_drop_geometry 

ggstatsplot::ggcorrmat(mdata_nogeo[,2:17])
```

# Building a non-spatial MLR

## Computing bw

```{r}
gwr_bw_train_ad <- bw.gwr(resale_price ~ floor_area_sqm + 
                            storey_order + remaining_lease_mths + 
                            PROX_CBD + PROX_ELDERLYCARE + PROX_HAWKER +
                            PROX_MRT + PROX_PARK + PROX_MALL + 
                            PROX_SUPERMARKET + WITHIN_350M_KINDERGARTEN +
                            WITHIN_350M_CHILDCARE + WITHIN_350M_BUS +
                            WITHIN_1KM_PRISCH,
                          data = train_data, 
                          approach="CV", 
                          kernel="gaussian", 
                          adaptive=TRUE, 
                          longlat = FALSE)
```

-   Data used is only Testing Data

## Model Calibration

```{r}
write_rds(gwr_bw_train_ad, "data/model/gwr_bw_train_ad.rds")
```

## Model Output

## Computing test data dw

-   (can skip and use gwr.predict, predictdata)

## Predicting with test data

```{r}
gw_pred <- gwr.predict(formula = resale_price ~ floor_area_sqm + 
                            storey_order + remaining_lease_mths + 
                            PROX_CBD + PROX_ELDERLYCARE + PROX_HAWKER +
                            PROX_MRT + PROX_PARK + PROX_MALL + 
                            PROX_SUPERMARKET + WITHIN_350M_KINDERGARTEN +
                            WITHIN_350M_CHILDCARE + WITHIN_350M_BUS +
                            WITHIN_1KM_PRISCH,
                          data = train_data,
                        predictdata = test_data, 
                       bw=20, 
                       kernel="gaussian", 
                       adaptive=TRUE, 
                       longlat=FALSE)
```

## Saving predicted values

```{r}
gwr_pred_df <- as.data.frame(gwr_pred$SDF$prediction) %>% 
  rename(gwr_pred = "gwr$SDF$prediction")
```

# Predictive Modelling with RF Method

## Data Preparation

We extract the coordinates of training and test data sets

```{r}
coords <- st_coordinates(HDB_sample)
coords_train <- st_coordinates(train_data)
coords_test <- st_coordinates(test_data)
```

Next, the code chunk below is used to drop the geometry column for both training and test datasets

```{r}
train_data_nogeom <- train_data %>% 
  st_drop_geometry()
```

## Calibrating RF Model

```{r}
set.seed(1234)
rf <- ranger(resale_price ~ floor_area_sqm + 
                            storey_order + remaining_lease_mths + 
                            PROX_CBD + PROX_ELDERLYCARE + PROX_HAWKER +
                            PROX_MRT + PROX_PARK + PROX_MALL + 
                            PROX_SUPERMARKET + WITHIN_350M_KINDERGARTEN +
                            WITHIN_350M_CHILDCARE + WITHIN_350M_BUS +
                            WITHIN_1KM_PRISCH,
             data=train_data_nogeom)
```

## Model Output Interpretation

-   \<insert text here\>

## Preparing the Test Data

```{r}
test_data_nogeom <- cbind(test_data, coords_test) %>% 
  st_drop_geometry()
```

## Predicting with rf

```{r}
rf_pred <- predict(rf, data=test_data_nogeom)
```

## Saving the predicted values

```{r}
rf_pred_df <- as.data.frame(rf_pred$predictions) %>% 
  rename(rf_pred = "rf_pred$predictions")
```

-   Take note of the naming convention

# Predictive Modelling with SpatialML 

## Determining Bandwidth

```{r}

```

## Calibrating with grf

```{r}
set.seed(1234)
grf_ad <- grf(formula = resale_price ~ floor_area_sqm + 
                            storey_order + remaining_lease_mths + 
                            PROX_CBD + PROX_ELDERLYCARE + PROX_HAWKER +
                            PROX_MRT + PROX_PARK + PROX_MALL + 
                            PROX_SUPERMARKET + WITHIN_350M_KINDERGARTEN +
                            WITHIN_350M_CHILDCARE + WITHIN_350M_BUS +
                            WITHIN_1KM_PRISCH,
              dframe=train_data_nogeom, 
              bw=20)
```

## Model Output

```{r}
grf_ad
```

## Preparing test data

```{r}
test_data_nogeom
```

## Predicting with test data

```{r}
grf_pred <- predict.grf(grf_ad, 
                        test_data_nogeom, 
                        x.var.name="X", 
                        y.var.name="Y", 
                        local.w = 1, 
                        global.w=0)
```

-   0 and 1 is a toggle between global and local model

## Saving predicted values

```{r}
grf_pred_df <- as.data.frame(grf_pred)
```

# Model Comparison 

## Combining Models' Output

```{r}
test_data_pred <- test_data %>% 
  select(resale_price) %>% 
  cbind(gwr_pred_df) %>% 
  cbind(rf_pred_df) %>% 
  cbind(grf_pred_df)
```

-   Each of this, has the same structure; header is the preparation name that is why the naming convention is as such

-   When appending back, it will be the same number of rows (in this case 500)

## Transposing Data

```{r}
test_longer <- test_data_pred %>% 
  st_drop_geometry() %>% 
  pivot_longer(cols = ends_with("pred"), 
               names_to = "model", 
               values_to = "predicted")
```

-   Transposition is only applied to grw_pred, rf_pred, grd_pred

## Renaming Variables 

```{r}
model_labels <- c(
  gwr_pred = "gwr", 
  rf_pred = "Random Forest", 
  grf_pred = "gwRF"
)

test_longer <- test_longer %>% 
  mutate(model = record(
    model, !!!model_labels
  ))
```

## Computing RMSE

```{r}
rmse_results <- test_longer %>% 
  group_by(model) %>% 
  rmse(trith = resale_price, 
       estimate = predicted) %>% 
  rename(rmse = .estimate) %>% 
  select(model, rmse)
```

# Model Comparison plots 

To prepare these plots, and to make the extent the same, you need to perform a `facet()`

## Scatter Plots 

## Bar Chart

```{r}
ggplot(rmse_results, 
       aes(x = reorder(mode, rmse, 
                       y=rmse,
                       fill="skyblue")) + 
         geom_bar(stat="identity", 
                  fill="skyblue", ))
```

## 
